\section{General Nested Expressions}

The following example is an expression, which we cannot compress with the previous theorem:
$$(a,b,c,d,e,abbcde \rightarrow bc, v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, (
    i,j,k,l \rightarrow iijkkl, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}
    ))$$
for $v^{(i)} \in \R^{d_{vi}}$ with $i \in [9]$, where $d_{vi} \in \N$ are appropriate dimensions.
This is because the output string $\bm{s_u} = iijkkl$ and the input string $\bm{\hat{s}_u} = abbde$ are not the same.
In the following, we will explore how to compress such expressions.
Note that, for the theorem, we use disjoint sets of symbols for the inner and outer expression.
This helps in the proof, and is not a real constraint in practice,
because we can just rename the symbols in different scopes.
For example, we could also write the above expression as
$$(ij, jjj \rightarrow i, A, (ik, kj \rightarrow iij, B, C)),$$
because the scope of each symbol does not reach into nested expressions,
and therefore the $j$ used in the outer expression and the $j$ used in the inner expression are treated as different symbols.

\begin{theorem}
    \label{thm:nested_einsum:general}
    For $i \in [m + n]$, let $T^{(i)}$ be an $n_i$-th order tensor with index string $\bm{s_i} \in S^{n_i}$.
    Let $\bm{s_u}$ be an index string for the $n_u$-th order tensor $U$, which is defined as follows:
    $$U := (\bm{s_{m + 1}},\dots,\bm{s_{m + n}} \rightarrow \bm{s_u}, T^{(m + 1)},\dots,T^{(m + n)})$$
    Also let $\bm{\hat{s}_u}$ be alternative index strings for $U$.

    Let $s_v$ be an index string and
    $$V := (\bm{s_1},\dots,\bm{s_m}, \bm{\hat{s}_u} \rightarrow \bm{s_v}, T^{(1)},\dots,T^{(m)}, U)$$
    where the first and second Einsum expression share no symbols.
    Then these nested Einsum expressions can also be compressed into a single Einsum expression.

    Let us clarify that the index string $\bm{s_u}$ corresponds to the output string of the inner expression,
    and the index string $\bm{\hat{s}_u}$ corresponds to the input string that is used for the input tensor $U$ in the outer expression.
    In our example, these are $\bm{s_u}$ and $\bm{\hat{s}_u}$:
    $$(a,b,c,d,e,\overbrace{abbcde}^{\bm{\hat{s}_u}} \rightarrow bc, v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, (
        i,j,k,l \rightarrow \overbrace{iijkkl}^{\bm{s_u}}, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}
        )).$$

    In contrast to \cref{thm:nested_einsum:simple}, we cannot just replace the input index string $\bm{\hat{s}_u}$ by all the input index strings in the inner Einsum expression $\bm{s_{m + 1}},\dots,\bm{s_{m + n}}$.
    Instead, we first need to apply a symbol map $\nu: S \rightarrow S$ to each of the index strings.
    This symbol map holds information about which symbols are effectively the used for the same index.

    For the definition of the map $\nu$, we first construct an undirected graph $G = (V, E)$ that we call \textit{symbol graph}.
    % $V = \sigma(\bm{s_v}) \cup \sigma(\bm{s_u}) \cup \sigma(\bm{\hat{s}_u}) \cup \bigcup_{i \in [m + n]} \sigma(\bm{s_i})$
    In the symbol graph, the nodes consist of all symbols from $\bm{s_u}$ and $\bm{\hat{s}_u}$.
    The edges are $E = \smallset{\smallset{s_{uj}, \hat{s}_{uj}} \mid j \in [n_u]}$,
    which connects all symbols from $\bm{s_u}$ and $\bm{\hat{s}_u}$ that share an index.
    The symbol graph for our example is illustrated in \cref{fig:nested_expressions:example_symbol_graph}.

    \begin{figure}[h]
        \centering
        \begin{tikzpicture}[node distance = 3cm, semithick]

            \node[state] (a)					{$a$};
            \node[state] (b) [right of=a]		{$b$};
            \node[state] (c) [right of=b] 		{$c$};
            \node[state] (d) [right of=c] 		{$d$};
            \node[state] (e) [right of=d] 		{$e$};
            \node (middle) at ($(a)!0.5!(b)$) {};
            \node[state] (i) [below of=middle] 		{$i$};
            \node[state] (j) [right of=i] 		{$j$};
            \node[state] (k) [right of=j] 		{$k$};
            \node[state] (l) [right of=k] 		{$l$};

            \node (x_center) at ($(a)!0.5!(j)$) {};
            \node (y_center) at ($(c)!0.5!(d|-k)$) {};
            \node (z_center) at ($(e)!0.5!(l)$) {};
            \node (x) [below of=x_center] {$x$};
            \node (y) [below of=y_center] {$y$};
            \node (z) [below of=z_center] {$z$};

            \path (a) edge (i);
            \path (i) edge (b);
            \path (b) edge (j);
            \path (c) edge (k);
            \path (k) edge (d);
            \path (e) edge (l);

            \begin{pgfonlayer}{background}
                \draw[gray!15, fill=gray!15, line width=12mm, line cap=round, line join=round] (a.center) -- (b.center) -- (j.center) -- (i.center) -- cycle;
                \draw[gray!15, fill=gray!15, line width=12mm, line cap=round, line join=round] (c.center) -- (d.center) -- (k.center) -- cycle;
                \draw[gray!15, fill=gray!15, line width=12mm, line cap=round, line join=round] (e.center) -- (l.center) -- cycle;
            \end{pgfonlayer}

        \end{tikzpicture}
        \caption{Symbol graph for the example}
        \label{fig:nested_expressions:example_symbol_graph}
    \end{figure}

    In the symbol graph, if two symbols are connected, then they are effectively the same index.
    Therefore, it makes sense assigning a symbol $s_C \in S \setminus V$ to each of the graphs components $C$.
    Then we can define $\nu$ as follows:
    $$\nu(s) := \begin{cases}
            s_C & \text{if $\exists C: \text{$C$ is a component of $G$ and $s \in C$}$} \\
            s   & \text{else}
        \end{cases}.$$
    In our example, the components are $\smallset{a,b,i,j}$, $\smallset{c,d,k}$, and $\smallset{e,l}$.
    Therefore we could use
    $$\nu(s) := \begin{cases}
            x & \text{if $s \in \smallset{a,b,i,j}$} \\
            y & \text{if $s \in \smallset{c,d,k}$}   \\
            z & \text{if $s \in \smallset{e,l}$}     \\
            s & \text{else}
        \end{cases}.$$

    The symbol map $\nu$ can be extended, such that it maps entire index strings instead of just symbols, by setting $\nu(\bm{s_i}) \in S^{n_i}, \nu(\bm{s_i})_j := \nu(s_{ij})$.
    Then we can write the substituted index strings by setting $\bm{\hat{s}_i} := \nu(\bm{s_i})$ for $i \in [m + n]$ and $\bm{\hat{s}_t} = \nu(\bm{s_t})$.
    With these index strings, the compressed Einsum expression is the following:
    $$V = (\bm{\hat{s}_1}, \dots, \bm{\hat{s}_{m + n}} \rightarrow \bm{\hat{s}_v}, T^{(1)},\dots,T^{(m + n)})$$
    which helps us to compress the example:
    \begin{gather*}
        (a,b,c,d,e,abbde \rightarrow bc, v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, (
        i,j,k,l \rightarrow iijkkl, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}
        ))\\
        =(x,x,y,y,z,x,x,y,z \rightarrow xy, v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}).
    \end{gather*}
\end{theorem}

\bigskip
For the proof of this theorem, we first need three lemmata, which essentially boil down to one intuitive thought:
The effective equality of two symbols can be expressed by multiplication with the unity matrix $\1_d$:
$$\left(\1_d\right)_{ij} := \begin{cases}
        \1 & \text{if $i = j$} \\
        \0 & \text{else}
    \end{cases},$$
for $i,j \in [d]$, where $\0$ and $\1$ indicate the neutral element of addition and multiplication in the given semiring respectively.

\input{chapters/nested/lemmata/lemma_1.tex}

This lemma intuitively means that we can replace any symbol in an index string of an input tensor of our choice with a new symbol by introducing the unity matrix with an appropriate index string as a factor.
Now the same holds for the index string of the output tensor $\bm{s_t}$, which will be the content of the next lemma.

\input{chapters/nested/lemmata/lemma_2.tex}

Now with these two lemmata, we can replace any symbol in any index string, regardless if it is and input string or the output string by introducing the unity matrix with an appropriate index string as a factor.
In the following lemma, we will show that any unity matrix factors can be removed again, by renaming certain symbols in all other index strings in the Einsum expression.

\input{chapters/nested/lemmata/lemma_3.tex}

From these three lemmata, \cref{thm:nested_einsum:general} follows with the following procedure:
\begin{enumerate}[label={\arabic*.}]
    \item Apply \cref{lemma:nested_einsum:1} to all of the symbols in the input string $\bm{\hat{s}_u}$ and therefore replace it with a new index string $\bm{s'_u} \in S^{n_u}$ where $\bm{s'_u}$ contains no duplicate symbols.
    \item Apply \cref{lemma:nested_einsum:2} to all of the symbols in the output string $\bm{s_u}$ and therefore replace it with the same new index string $\bm{s'_u}$.
    \item Apply \cref{thm:nested_einsum:simple} to the nested expression and therefore compress the nested expression into a single expression with lots of unity matrices.
          This is possible because the input string and output string for $U$ are both $\bm{s'_u}$ now.
    \item Apply \cref{lemma:nested_einsum:3} to remove unity matrices from the compressed expression until there are no more of those unity matrices left, which were introduced in step one and two.
\end{enumerate}

\begin{proof}
    \small
    For the proof, we again demonstrate the arguments on the example used in \cref{thm:nested_einsum:general} for better understanding:
    $$(a,b,c,d,e,abbcde \rightarrow bc,v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, (
        i,j,k,l \rightarrow iijkkl, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}
        ))$$
    Applying step one and two to this example results in the following Einsum expression if $\bm{s'_u} = s_1 s_2 s_3 s_4 s_5 s_6$:
    \begin{gather*}
        (a,b,c,d,e,s_1 s_2 s_3 s_4 s_5 s_6, a s_1, b s_2, b s_3, c s_4, d s_5, e s_6 \rightarrow bc,\\
        v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, \1, \1, \1, \1, \1, \1,\\
        (i,j,k,l, i s_1, i s_2, j s_3, k s_4, k s_5, l s_6 \rightarrow s_1 s_2 s_3 s_4 s_5 s_6,\\
        v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}, \1, \1, \1, \1, \1, \1)),
    \end{gather*}
    where we used $\1$ to indicate unity matrices of different sizes, because the sizes can be derived from the context and are not important for better understanding.
    Applying step three results in the following compressed expression:
    \begin{gather*}
        (a,b,c,d,e,i,j,k,l, i s_1, i s_2, j s_3, k s_4, k s_5, l s_6, a s_1, b s_2, b s_3, c s_4, d s_5, e s_6 \rightarrow bc,\\
        v^{(1)}, v^{(2)}, v^{(3)}, v^{(4)}, v^{(5)}, v^{(6)}, v^{(7)}, v^{(8)}, v^{(9)}, \1, \1, \1, \1, \1, \1, \1, \1, \1, \1, \1, \1)
    \end{gather*}

    The only fact that remains to be understood is why the removal the unity matrices in step four leads to the transformation described in \cref{thm:nested_einsum:general}.
    For this, we construct another undirected graph $G' = (V', E')$ that we call \textit{extended symbol graph}.
    In the symbol graph, the nodes consist of all symbols from $\bm{s_u}$, $\bm{s'_u}$, and $\bm{\hat{s}_u}$.
    An edge $\smallset{u,v}$ exists precisely when the compressed expression contains a unity matrix with index string $uv$, that was introduced in step one or step two.
    The extended symbol graph for our example is illustrated in \cref{fig:nested_expressions:example_extended_symbol_graph}.

    \begin{figure}[h]
        \centering
        \begin{tikzpicture}[node distance = 3cm, semithick]

            \node[state] (a) at (0, 0)		{$a$};
            \node[state] (b) at (3, 0)		{$b$};
            \node[state] (c) at (6, 0) 		{$c$};
            \node[state] (d) at (9, 0) 		{$d$};
            \node[state] (e) at (12, 0) 	{$e$};

            \node[state] (s1) at (-0.25, -2)        {$s_1$};
            \node[state] (s2) at (2.25, -2)         {$s_2$};
            \node[state] (s3) at (4.75, -2)         {$s_3$};
            \node[state] (s4) at (7.25, -2)         {$s_4$};
            \node[state] (s5) at (9.75, -2)         {$s_5$};
            \node[state] (s6) at (12.25, -2)        {$s_6$};

            \node[state] (i) at (1.5, -4) 		{$i$};
            \node[state] (j) at (4.5, -4) 		{$j$};
            \node[state] (k) at (7.5, -4) 		{$k$};
            \node[state] (l) at (10.5, -4) 		{$l$};

            \path (a) edge (s1);
            \path (b) edge (s2);
            \path (b) edge (s3);
            \path (c) edge (s4);
            \path (d) edge (s5);
            \path (e) edge (s6);

            \path (i) edge (s1);
            \path (i) edge (s2);
            \path (j) edge (s3);
            \path (k) edge (s4);
            \path (k) edge (s5);
            \path (l) edge (s6);

        \end{tikzpicture}
        \caption{Extended symbol graph for the example}
        \label{fig:nested_expressions:example_extended_symbol_graph}
    \end{figure}

    Now, every newly introduced unity matrix is represented by an edge in the extended symbol graph.
    If a unity matrix is removed with \cref{lemma:nested_einsum:3}, the symbols connected by the representing edge collapse into one symbol,
    and the rest of the extended symbol graph stays the same.
    An example of this collapse is illustrated in \cref{fig:nested_expressions:collapsed_extended_symbol_graph}.

    \begin{figure}[h]
        \centering
        \begin{subfigure}{0.49\textwidth}
            \begin{tikzpicture}[node distance = 3cm, semithick]

                \node[state] (a) at (0, 0)		{$a$};
                \node[state] (b) at (3, 0)		{$b$};

                \node[state] (s1) at (-0.25, -2)        {$s_1$};
                \node[state] (s2) at (2.25, -2)         {$s_2$};
                \node[state] (s3) at (4.75, -2)         {$s_3$};

                \node[state] (i) at (1.5, -4) 		{$i$};
                \node[state] (j) at (4.5, -4) 		{$j$};

                \path (a) edge (s1);
                \path[red] (b) edge (s2);
                \path (b) edge (s3);

                \path (i) edge (s1);
                \path (i) edge (s2);
                \path (j) edge (s3);

            \end{tikzpicture}
        \end{subfigure}%
        \begin{subfigure}{0.49\textwidth}
            \begin{tikzpicture}[node distance = 3cm, semithick]

                \node[state] (a) at (0, 0)		{$a$};

                \node[state] (s1) at (-0.25, -2)    {$s_1$};
                \node[state, red] (x) at (2.25, -2)         {$x$};
                \node[state] (s3) at (4.75, -2)         {$s_3$};

                \node[state] (i) at (1.5, -4) 		{$i$};
                \node[state] (j) at (4.5, -4) 		{$j$};

                \path (a) edge (s1);
                \path (x) edge (s3);

                \path (i) edge (s1);
                \path (i) edge (x);
                \path (j) edge (s3);

            \end{tikzpicture}
        \end{subfigure}
        \caption{First component of the extended symbol graph after removing the unity matrix represented by the edge $\smallset{b, s_2}$}
        \label{fig:nested_expressions:collapsed_extended_symbol_graph}
    \end{figure}

    Therefore, after repeatedly applying \cref{lemma:nested_einsum:3},
    all nodes that are part of the same component in the extended symbol graph will collapse into one symbol.
    Now the only thing left to show is that two nodes are connected in $G$ exactly if they are also connected in $G'$.
    This can be seen by understanding the edges of $G$ and $G'$.
    In $G$, two symbols $s_{ui}$ and $\hat{s}_{uj}$ share an edge if they share a position $i = j$.
    In $G'$, two symbols $s_{ui}$ and $\hat{s}_{uj}$ share a neighbour $s'_{uk}$ if both were replaced by $s'_{uk}$ in step one and step two,
    which happens exactly if they share a position $i = j = k$.
    Therefore every edge in $G$ is represented by a shared neighbour in $G'$.
    Now, because every $s'_{uk}$ has exactly two neighbours, and because there are no direct edges between any symbols of $\bm{s_u}$ and $\bm{\hat{s}_u}$, there are no more edges in $G'$ other than the ones that contribute to a shared neighbour $s'_{uk}$.
    Then, because sharing an edge in $G$ is the same as sharing a neighbour in $G'$, two symbols are in the same component in $G$ precisely when they are also in the same component in $G'$.

    Therefore collapsing every edge in the extended symbol graph leads to the symbol map defined in \cref{thm:nested_einsum:general}.
\end{proof}