\section{Introducing Duplications}

The following example is an expression, which we cannot compress with the previous theorem:
$$(ij, jjj \rightarrow i, A, (kl, lo \rightarrow kko, B, C))$$
for $A \in \R^{a \times b}, B \in \R^{b \times c}, C \in \R^{c \times b}$.
In the following theorem, we will explore how to compress expressions such as this one.
Note that, for the theorem, we use disjoined sets of symbols for the inner and outer expression.
This helps in the proof, and is not a real constraint in practice,
because we can just rename the symbols in different scopes.
For example, we could also write the above expression as
$$(ij, jjj \rightarrow i, A, (ik, kj \rightarrow iij, B, C)).$$

\begin{theorem}
    \label{thm:nested_einsum:2}

    For $i \in [m + n]$, let $T^{(i)}$ be an $n_i$-th order tensor with index strings $\bm{s_i} \in S^{n_i}$.
    Let $\bm{s_u}$ be an index string for the $n_u$-th order tensor $U$, which is defined as follows:
    $$U := (\bm{s_{m + 1}},\dots,\bm{s_{m + n}} \rightarrow \bm{s_u}, T^{(m + 1)},\dots,T^{(m + n)})$$
    Also let $\bm{\hat{s}_u}$ be alternative index strings for $U$ with $s_{uj} = s_{uj'} \implies \hat{s}_{uj} = \hat{s}_{uj'}$ for all $j, j' \in [n_u]$,
    which means that $\bm{\hat{s}_u}$ can only introduce new symbol duplications, and cannot remove any.

    In our example, $\bm{s_u} = kko$ and $\bm{\hat{s}_u} = jjj$.
    This does not break the symbol duplication of the first and second index,
    and introduces a new duplication on the third index.

    Let $s_v$ be an index string and
    $$V := (\bm{s_1},\dots,\bm{s_m}, \bm{\hat{s}_u} \rightarrow \bm{s_v}, T^{(1)},\dots,T^{(m)}, U)$$
    such that the first and second Einsum expression share no symbols.
    Then these nested Einsum expressions can also be compressed into a single Einsum expression.

    In contrast to \autoref{thm:nested_einsum:1}, we cannot just replace the input index string $\bm{\hat{s}_u}$ by all the input index strings in the inner Einsum expression $\bm{s_{m + 1}},\dots,\bm{s_{m + n}}$.
    Instead, we first need to apply a symbol map.
    Let $\nu: S \rightarrow S$ such that
    $$\nu(s) := \begin{cases}
            \hat{s}_{uj} & \text{if }\exists j \in [n_u]: s_{uj} = s \\
            s            & \text{else}
        \end{cases}$$
    which maps symbols in $\bm{s_u}$ to the symbol at the same index in $\bm{\hat{s}_u}$ and all other symbols to themselves.

    This symbol map holds information about which symbols will be iterated over at the same time in the outer expression.
    In our example, the interesting parts of the map are $\nu(k) = j$ and $\nu(o) = j$, which means that $k$ and $j$ will be iterated over at the same time.

    $\nu$ can be extended, such that it maps entire index strings instead of just symbols, by setting $\nu(\bm{s_i}) \in S^{n_i}, \nu(\bm{s_i})_j := \nu(s_{ij})$.
    Then we can write the substituted index strings by setting $\bm{\hat{s}_i} := \nu(\bm{s_i})$ for $i \in [m + 1, m + n]$.

    Then the compressed Einsum expression is the following:
    $$V = (\bm{s_1},\dots,\bm{s_m}, \bm{\hat{s}_{m + 1}}, \dots, \bm{\hat{s}_{m + n}} \rightarrow \bm{s_v}, T^{(1)},\dots,T^{(m + n)})$$
    which helps us to compress the example:
    $$(ij, jjj \rightarrow i, A, (kl, lo \rightarrow kko, B, C)) = (ij, jl, lj \rightarrow i, A, B, C)$$
\end{theorem}

\begin{proof}
    \small
    The fundamental idea behind this theorem is, that by using the index string $\bm{\hat{s}_u}$, we only iterate over a sub-space of the indices that we defined for the computation of $U$.
    To formulate this, we need some idea of which multi-indices we iterate over.
    Therefore, let $\mathcal{M}:\bm{s} := \smallset{M: \bm{s} \mid M \in \mathcal{M}}$ for an index string $\bm{s}$ and a multi-index space $\mathcal{M}$.

    Let $F, F', B, B'$ be the free and bound symbols of the second (outer) and first (inner) einsum expression respectively.
    W.l.o.g. they are all non-empty.
    From them we can derive $\mathcal{F}, \mathcal{F}', \mathcal{B}, \mathcal{B}'$ as in the definition.
    Then $(\mathcal{F} \times \mathcal{B}):\bm{\hat{s}_u} \subseteq \mathcal{F}':\bm{s_u}$.
    This follows from $d_{s_{uj}} = d_{\hat{s}_{uj}}$ for $j \in [n_u]$,
    and the amount of symbols in the projection of $(\mathcal{F} \times \mathcal{B}):\bm{\hat{s}_u}$ being smaller or equal to the amount of symbols in the projection of $\mathcal{F}':\bm{s_u}$.
    The first fact is true per the definition of einsum.
    The second fact can be rewritten as $\abs{\sigma(\bm{\hat{s}_u})} \leq \abs{\sigma(\bm{s_u})}$ and follows directly from the constraint $s_{uj} = s_{uj'} \implies \hat{s}_{uj} = \hat{s}_{uj'}$ for all $j,j' \in [n_u]$.

    Then
    $$\forall \bm{f'} \in \mathcal{F}': U_{\bm{f'}: \bm{s_u}} = \bigoplus\limits_{\bm{b'} \in \mathcal{B}'}\bigodot\limits_{i = m + 1}^{m + n} T^{(i)}_{(\bm{f'}, \bm{b'}):\bm{s_{i}}}$$
    and therefore
    $$\forall (\bm{f}, \bm{b}) \in \mathcal{F} \times \mathcal{B}: U_{(\bm{f}, \bm{b}): \bm{\hat{s}_u}} = \bigoplus\limits_{\bm{b'} \in \mathcal{B}'}\bigodot\limits_{i = m + 1}^{m + n} T^{(i)}_{(\bm{f}, \bm{b}, \bm{b'}):\bm{\hat{s}_{i}}}$$
    because of the previous observation,
    and because the bound symbols of the expression, which are used in $\bm{b'}$, do not occur in $\bm{s_u}$, and are therefore not changed by the symbol map $\nu$.

    Therefore
    \begin{align*}
        V                                                         & = (\bm{s_1},\dots,\bm{s_m}, \bm{\hat{s}_u} \rightarrow \bm{s_v}, T^{(1)},\dots,T^{(m)}, U)                                                                                                                                                                \\
        \iff \forall \bm{f} \in \mathcal{F}: V_{\bm{f}: \bm{s_v}} & = \bigoplus\limits_{\bm{b} \in \mathcal{B}} \bigodot\limits_{i = 1}^{m} T^{(i)}_{(\bm{f}, \bm{b}):\bm{s_i}} \odot U_{(\bm{f}, \bm{b}):\bm{\hat{s}_u}}                                                                                                     \\
                                                                  & = \bigoplus\limits_{\bm{b} \in \mathcal{B}} \bigodot\limits_{i = 1}^{m} T^{(i)}_{(\bm{f}, \bm{b}):\bm{s_i}} \odot \bigoplus\limits_{\bm{b'} \in \mathcal{B}'} \bigodot\limits_{i' = m + 1}^{m + n} T^{(i')}_{(\bm{f}, \bm{b}, \bm{b'}):\bm{\hat{s}_{i'}}} \\
                                                                  & = \bigoplus\limits_{\bm{b} \in \mathcal{B} \times \mathcal{B}'} \bigodot\limits_{i = 1}^{m} T^{(i)}_{(\bm{f}, \bm{b}):\bm{s_i}} \odot \bigodot\limits_{i = m + 1}^{m + n} T^{(i)}_{(\bm{f}, \bm{b}):\bm{\hat{s}_i}}                                       \\
        \iff V                                                    & = (\bm{s_1}, \dots, \bm{s_m}, \bm{\hat{s}_{m + 1}}, \dots, \bm{\hat{s}_{m + n}} \rightarrow \bm{s_v}, T^{(1)}, \dots, T^{(m + n)})
    \end{align*}
    where the third equality holds because we only iterate over a sub-space of the indices that we defined for the computation of $U$,
    and because the first and second einsum expression share no symbols.
    The rest of the steps are the same as in \autoref{thm:nested_einsum:1}.
\end{proof}
\bigskip